---
title: "Practicum 1"
output: html_notebook
author: 
  - Nickhil Tekwani, tekwani.n@northeastern.edu
  - Anurag Arasan, arasan.a@northeastern.edu
---

# Part 1: DB Connection to AWS MYSQL Instance
```{r}
library(RMySQL)
library(sqldf)
options(sqldf.driver = "SQLite")

db_user <- "cs5200"
db_password <- "databases2022"
db_name <- "Practicum1_DB"
db_host <- "cs5200-dbs.cejiubjir362.us-east-2.rds.amazonaws.com" # AWS Host
db_port <- 3306
dbcon <- dbConnect(MySQL(),
  user = db_user, password = db_password,
  dbname = db_name, host = db_host, port = db_port
)
```

## Part 1.5: Set Up Schema for DB
```{sql connection=dbcon}
DROP TABLE IF EXISTS incidents
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS airports
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS conditions
```

```{sql connection=dbcon}
CREATE TABLE conditions(
  cid INTEGER PRIMARY KEY AUTO_INCREMENT,
  `condition` TEXT NOT NULL,
  explanation TEXT
  )
```

```{sql connection=dbcon}
CREATE TABLE airports(
  aid INTEGER PRIMARY KEY AUTO_INCREMENT, 
  airportName TEXT NOT NULL, 
  airportCode TEXT, 
  state TEXT NOT NULL
  )
```

```{sql connection=dbcon}
CREATE TABLE incidents(
rid INTEGER NOT NULL PRIMARY KEY,
date DATE NOT NULL,
origin INTEGER,
airline TEXT NOT NULL,
aircraft TEXT NOT NULL,
flightPhase ENUM("Takeoff", "Landing", "Inflight", "Unknown") NOT NULL,
altitude INTEGER NOT NULL,
conditions INTEGER,
warning ENUM("Y", "N") NOT NULL,
FOREIGN KEY (conditions) REFERENCES conditions(cid),
FOREIGN KEY (origin) REFERENCES airports(aid)
)
```

# Part 2: Load in Bird Strikes CSV and pre-process data accordingly
```{r}

# load csv
df <- read.csv("BirdStrikesData-V2.csv",
  header = TRUE, stringsAsFactors = FALSE
)

# harmonize flight phases
df <- sqldf(c("UPDATE df
              SET flight_phase =
              CASE
                WHEN flight_phase = 'Climb' THEN 'Takeoff'
                WHEN flight_phase = 'Landing Roll' THEN 'Landing'
                WHEN flight_phase = 'Approach' THEN 'Landing'
                WHEN flight_phase = 'Take-off run' THEN 'Takeoff'
                WHEN flight_phase = 'Descent' THEN 'Landing'
                Else 'Unknown'
              END;", "SELECT * FROM df"))

# remove all military flights
df <- sqldf(c("DELETE FROM df WHERE airline = 'MILITARY'", "SELECT * FROM df"))

conditions_df <- sqldf(c("CREATE TABLE temp(cid INTEGER PRIMARY KEY AUTOINCREMENT,
  condition TEXT)", "INSERT INTO temp(condition)
  SELECT DISTINCT
   sky_conditions AS condition
   FROM df", "SELECT * FROM temp"))

airports_df <- sqldf(c(
  "CREATE TABLE temp(aid INTEGER PRIMARY KEY AUTOINCREMENT,
  airportName TEXT, airportCode TEXT, state TEXT)",
  "INSERT INTO temp(airportName, state)
  SELECT DISTINCT
   airport,
   origin
   FROM df", "SELECT * FROM temp"
))
#
df <- sqldf("SELECT rid,airport,flight_date,airline,aircraft,aid AS origin,flight_phase,altitude_ft, pilot_warned_flag,sky_conditions FROM df
JOIN airports_df ON df.airport = airports_df.airportName AND df.origin = airports_df.state")
#
incidents_df <- sqldf("SELECT
   rid,
   flight_date AS date,
   airline,
   aircraft AS aircraft,
   sky_conditions AS conditions,
   origin,
   flight_phase AS flightPhase,
   altitude_ft AS altitude,
   pilot_warned_flag AS warning
   FROM df")

incidents_df["date"] <- as.Date(incidents_df$date, "%m/%d/%y")

incidents_df <- sqldf("SELECT rid,date,airline,aircraft,origin,flightPhase,altitude, warning,cid
AS conditions FROM incidents_df
JOIN conditions_df ON incidents_df.conditions = conditions_df.condition")

incidents_df <- sqldf(c(
  "UPDATE incidents_df SET airline = 'UNKNOWN' WHERE airline = ''",
  "SELECT * FROM incidents_df"
))

incidents_df <- sqldf(c(
  "UPDATE incidents_df SET aircraft = 'Unknown' WHERE aircraft = ''",
  "SELECT * FROM incidents_df"
))
airports_df <- sqldf(c(
  "UPDATE airports_df SET airportName = 'Unknown' WHERE airportName = ''",
  "SELECT * FROM airports_df"
))
airports_df
```

```{r}
incidents_df
```

# Part 3: Load data into tables and show that it loaded appropriately
```{r}
dbWriteTable(dbcon, "conditions", conditions_df, row.names = FALSE, append = TRUE)
```

```{sql connection=dbcon}
# SHOW THAT CONDITIONS DATA INSERTED CORRECTLY
SELECT * FROM conditions
```

```{r}
dbWriteTable(dbcon, "airports", airports_df, row.names = FALSE, append = TRUE)
```
```{sql connection=dbcon}
# SHOW THAT AIRPORTS DATA INSERTED CORRECTLY
SELECT * FROM airports
```

```{r}
dbWriteTable(dbcon, "incidents", incidents_df, row.names = FALSE, append = TRUE)
```
```{sql connection=dbcon}
# SHOW THAT INCIDENTS DATA INSERTED CORRECTLY
SELECT * FROM incidents
```


# Part 4: Find the top-10 airlines with the most number of incidents
```{sql connection=dbcon}
SELECT airline FROM incidents
GROUP BY airline
ORDER BY COUNT(*) DESC
LIMIT 10
```

# Part 5: Find the flight phase(s) that had an above average number bird strike incidents
Note that the average is calculated by taking a count of all incidents, then dividing by the number of unique flightPhases that exist.
```{sql connection=dbcon}
SELECT flightPhase, COUNT(*) AS numberOfIncidents FROM incidents
GROUP BY flightPhase 
HAVING numberOfIncidents > (SELECT COUNT(*)/COUNT(DISTINCT flightPhase) AS avgIncidents FROM incidents)
```

# Part 6: Find the month with the most number of bird strike incidents
```{sql connection=dbcon}
SELECT MONTHNAME(date) AS month, COUNT(*) AS numOfBirdStrikeIncidents FROM incidents
GROUP BY MONTH(date)
ORDER BY COUNT(*) DESC
LIMIT 1
```

# Part 7: Create a stored procedure that adds a new incident to the database
```{sql connection=dbcon}
DROP PROCEDURE IF EXISTS AddIncident
```

We assume that anyone using this procedure is aware of the data constraints of the schema and will give inputs formatted as such.
```{sql connection=dbcon}

CREATE PROCEDURE AddIncident(
  IN rid TEXT,
  IN formattedDate TEXT,
  IN aircraft TEXT,
  IN airline TEXT,
  IN warning TEXT,
  IN flightPhase TEXT,
  IN altitude INTEGER,
  IN airport TEXT,
  IN cond TEXT,
  IN state TEXT)
  
BEGIN

  DECLARE tempCond INTEGER;
  DECLARE tempAirport INTEGER;
  DECLARE tempCid INTEGER;
  DECLARE tempAid INTEGER;
  
  SELECT COUNT(cid) INTO tempCond FROM conditions WHERE conditions.condition = cond;
  IF tempCond = 0 THEN
    INSERT INTO conditions (`condition`) VALUES (cond);
  END IF;
    
  
  SELECT COUNT(aid) INTO tempAirport FROM airports WHERE airports.airportName = airport;
  IF tempAirport = 0 THEN
    INSERT INTO airports (airportName, state) VALUES (airport, state);
  END IF;
    
  
  SELECT cid INTO tempCid FROM conditions WHERE conditions.condition = cond;

  SELECT aid INTO tempAid FROM airports WHERE airports.airportName = airport AND airports.state = state;
  
  
  
  INSERT INTO incidents (rid, date, airline, aircraft, flightPhase, altitude, warning, origin, conditions)
  VALUES (rid, formattedDate, airline, aircraft, flightPhase, altitude, warning, tempAid, tempCid);

  
END;
```


To check if the stored procedure works correctly, we give a list of correct input parameters for adding to the incidents table.
```{sql connection=dbcon}
CALL AddIncident(100, "2022-06-06", "Airplane", "American Airlines", "Y", "Landing", 200, "SFO International Airport", "Overcast", "California")
```

Showing that the new record was added to incidents table correctly. We can see that the new record for rid 100 is added, and also that a new aid for a new airport was added (1140 on origin column). The procedure also successfully looked up the cid for the given warning ("Overcast") and was shown accordingly in conditions column (3).
```{sql connection=dbcon}
SELECT * FROM incidents WHERE rid = 100;
```

We can also dynamically add a new condition. As shown below:
```{sql connection=dbcon}
CALL AddIncident(101, "2022-06-06", "Airplane", "American Airlines", "Y", "Landing", 200, "SFO International Airport", "New Condition Overcast", "California")
```
The following record with rid 101 has a condition code of 4, which did not previously exist. 
```{sql connection=dbcon}
SELECT * FROM incidents WHERE rid = 101;
```

Close connection to DB
```{r}
dbDisconnect(dbcon)
```




